Supplementary figures\nCapturing regeneration in models of forest dynamics

### 2023-04-14 manuscript/manuscriptRW.Rmd
changes based on PICUS team feedback

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g. in
the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Some models use calibration
against local canopy-level data to define appropriate ingrowth levels,
which is likely to strongly hamper the general applicability of these
models e.g. under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species diversity due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@schumacher2006] or by restricting the number of seeds per species in
the seedbank [@lischke2006]. However, simulated species composition is
usually exceedingly sensitive to that level, whereas the parameters of
such functions are poorly constrained by field data. Lastly, there is
often a problem with insufficient observational constraints on parameter
values for models that start from very small tree sizes (e.g., 10 cm of
height or even seeds) and track tree development in a process-oriented
manner by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

@@

Fifteen models of forest dynamics (including two models featuring
variants) were used to assess simulated tree establishment. They
simulate forest dynamics at the stand, landscape or global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SIBYLA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment processes.

@@

The starting point for tree establishment in the models differs as well,
ranging from seedbank, seed or seedling to sapling (i.e., trees much
larger than 10 cm height, often ca. 2-3 m tall). Models that start from
seed need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that start
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This later approach
reduces model complexity but comes at the cost of blurring process
representation.

In the models used in this study, the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table \@ref(tab:tableModels)). Two
models, ForClim and ForCEEPS, were used here with two alternative
variants of establishment. ForClim variant 1 [@bugmann1996] is based on
a recruitment module that adheres closely to the concept introduced by
@botkin1972, whereas ForClim variant 11 is adopting a different approach
where individual species properties and their relation to the
environment are important [@huber2020]. These two model variants allow
us to evaluate a more process-based and complex module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e., simulated actual
composition and relative abundance of species in the plot) via seed
trees for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth
thresholds for the sites differed between 0 and 10 cm. For our study, we
defined two data sets where one included 165 sites with a diameter
threshold of 7 cm or lower, and the other included another 35 sites with
diameter thresholds between 7 and 10 cm. Thus, we focused on two data
sets with harmonized diameter thresholds of 7 cm and 10 cm,
respectively.

Plot size ranged from 0.02 ha to 5.52 ha, and the data were further
processed and aggregated following @käber2022 to provide ingrowth rates
per ha and per decade. The observations used in this study featured
30,900 established trees. Ingrowth rates per site, sample and
decade ranged from zero to 1246 trees per ha, with a mean of 56 trees
per ha. Adult species composition was also available for the simulated
species. For more details on this unique dataset and the detailed
establishment information, cf. @käber2022.

@@

```{r combinedRichness, echo=FALSE, fig.cap="Mean Shannon index across all samples per site for observed and vs.simulated data. n indicates the number of models falling in each group. A) Examples for the three trends across models for species diversity at the stand level. B) Examples for the three trends across models for ingrowth at the 7 cm threshold. The full data are shown in Figures S5 and S6.", out.width='75%'}
knitr::include_graphics(here::here("figures", "combined_richness.png"))
```

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall species diversity levels in the
ingrowth were well within the observed range for half of the models (7
out of 14). Thus, while most of the models are facing difficulties to
quantitatively match ingrowth levels (cf. above), their performance is
better when it comes to matching the composition of ingrowth as a
function of abiotic and biotic conditions. Most models maintained or
even decreased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences
were not significant. Diversity variations may be explained by the way
the individual models consider establishment processes
[@konig_tree_2022], e.g., in the absence of adults, ingrowth can be
linked to either seed rain from dispersal or a seed/seedling bank;
neither process is incororated across many models studied here, however.

@@

Competition for light as a strong filter for tree establishment has been
widely documented [@messier1999; @collet2006; @berdanier2016], but the
models examined here did not reproduce this expectation. However, it is
difficult to measure light availability at large spatial and temporal
scales. We used total stand basal area as a proxy for light availability
[cf. @schmid2021]. We found pronounced differences in the stand basal
area ranges simulated by the models, but also between the models and the
observations. This made it impossible to evaluate the establishment for
the extremes of the stand density ranges in some models. For example,
ingrowth levels at low stand densities are relevant to assess how well
forests are recovering e.g. after gap creation due to disturbance
[@seidl2022; @grubb1977]. At the level of the simulated one-hectare
samples, average basal area is typically not very low as long as no
larger disturbances occur, which was explicitly excluded in our
simulation protocol. Towards the other end of the spectrum, i.e., with
increasing stand basal area, it would be reasonable to expect that the
ingrowth of the different species would become sparser and drop out
entirely under low-light conditions [@klopcic2012; @zell2019]. However,
few models showed this trend, thus indicating that the relation between
ingrowth and light availability is not yet captured correctly in most
models. Yet, several of the models that did feature an increase of
ingrowth with increasing basal area include a feedback between seed
production of mature trees and ingrowth. Thus, it seems that higher
light competition does not sufficiently compensate for increased seed
availability with higher basal area due to more mature trees.

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021, they found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This, may at
least partly explain e.g. the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

Based on the results of this study, it is clear that models of forest
dynamics need a focus on tree establishment modules to complement
earlier efforts to make them more robust regarding growth and mortality.
It remain uncertain what level of detail is required to model
establishment processes, and this must be addressed in future research.
We recommend that the further development of the establishment modules
is implemented as additional features that can be traced back, as done
here for the variants of ForClim and ForCEEPS, and that model complexity
and structure must always be connected with modeling objectives
[@albrich2020]. If it should be necessary to include more detail in the
establishment models, this will come with higher parameterisation
efforts. This will most likely lead to lower generalization because the
required data will have to be collected from specific locations, as
currently there is no general establishment dataset available.

>>>>

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g. in
the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Some models use calibration
against local canopy-level data to define appropriate ingrowth levels,
which is likely to strongly hamper the general applicability of these
models e.g. under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species diversity due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@schumacher2006] or by restricting the number of seeds per species in
the seedbank [@lischke2006]. However, simulated species composition is
usually exceedingly sensitive to such a formulations, whereas the
parameters of such functions are poorly constrained by field data.
Lastly, there is often a problem with insufficient observational
constraints on parameter values for models that start from very small
tree sizes (e.g., 10 cm of height or even seeds) and track tree
development in a process-oriented manner by considering a multitude of
ecological influences, rather than emphasizing tree establishment into a
larger size class [e.g., @wehrli2006].

@@

Fifteen models of forest dynamics (including two models featuring model
variants) were used to assess simulated tree establishment. They
simulate forest dynamics at the stand, landscape or global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SIBYLA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment processes.

@@

The starting point for tree establishment in the models differs as well,
ranging from seedbank, seed or seedling to sapling (i.e., trees much
larger than 10 cm height, often ca. 2-3 m tall). Models that start from
seed need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that start
from saplings have to aggregate via parameterizations several ecological
processes which are not treated explicitly. This later approach reduces
model complexity but comes at the cost of blurring process
representation.

In the models used in this study, the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity across
all processes \>0, Table \@ref(tab:tableModels)). Two models, ForClim
and ForCEEPS, were used here with two alternative variants of
establishment. ForClim variant 1 [@bugmann1996] is based on a
recruitment module that adheres closely to the concept introduced by
@botkin1972, whereas ForClim variant 11 is adopting a different approach
where individual species properties and their relation to the
environment are important [@huber2020]. These two model variants allow
us to evaluate a more process-based and complex module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e., simulated actual
composition and relative abundance of species in the plot) via seed
trees for the quantity and quality (e.g., diversity and composition) of
simulated ingrowth, as one variant includes this feedback whereas the
other does not.

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, climatic
water balance, soil quality and exposition (i.e., slope and aspect).
Each of these sites featured at least two consecutive measurements.
Ingrowth thresholds for the sites differed between diameters of 0 and 10
cm. For our study, we defined two data sets where one included 165 sites
with a diameter threshold of 7 cm or lower, and the other included
another 35 sites with diameter thresholds between 7 and 10 cm.

Plot size ranged from 0.02 ha to 5.52 ha, and the data were further
processed and aggregated following @käber2022 to provide ingrowth rates
per ha and per decade. The observations used in this study featured
30,900 established trees. Ingrowth rates per site, sample and
decade ranged from zero to 1246 trees per ha, with a mean of 56 trees
per ha. Adult species composition was also available at the plots. For
more details on this unique dataset and the detailed establishment
information, cf. @käber2022.

@@

```{r combinedRichness, echo=FALSE, fig.cap="Mean Shannon index across all samples per site for observed and simulated data. n indicates the number of models falling in each group. A) Examples for the three trends across models for species diversity at the stand level. B) Examples for the three trends across models for ingrowth at the 7 cm threshold. The full data are shown in Figures S3 and S4.", out.width='75%'}
knitr::include_graphics(here::here("figures", "combined_richness.png"))
```

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall species diversity levels in the
ingrowth were well within the observed range for half of the models (7
out of 14). Thus, while most of the models are facing difficulties to
quantitatively match ingrowth levels (cf. above), their performance is
better when it comes to matching the composition of ingrowth as a
function of abiotic and biotic conditions. Most models maintained or
even decreased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences
were not significant. Diversity variations may be explained by the way
the individual models consider establishment processes
[@konig_tree_2022], e.g., in the absence of adults, ingrowth can be
linked to either seed rain from dispersal or a seed/seedling bank;
neither process is incorporated across many models studied here,
however.

@@

Competition for light as a strong filter for tree establishment has been
widely documented [@messier1999; @collet2006; @berdanier2016], but the
models examined here did not reproduce this expectation. However, it is
difficult to measure light availability at large spatial and temporal
scales. We used total stand basal area as a proxy for light availability
[cf. @schmid2021]. However we could not consider light availability
restrictions caused by ground vegetation which might be an important
filter for some models [@woltjer2008]. We found pronounced differences
in the stand basal area ranges simulated by the models, but also between
the models and the observations. This made it impossible to evaluate the
establishment for the extremes of the stand density ranges in some
models. For example, ingrowth levels at low stand densities are relevant
to assess how well forests are recovering e.g. after gap creation due to
disturbance [@seidl2022; @grubb1977]. At the level of the simulated
one-hectare samples, average basal area is typically not very low as
long as no larger disturbances occur, which was explicitly excluded in
our simulation protocol. Towards the other end of the spectrum, i.e.,
with increasing stand basal area, it would be reasonable to expect that
the ingrowth of the different species would become sparser and drop out
entirely under low-light conditions [@klopcic2012; @zell2019]. However,
few models showed this trend, thus indicating that the relation between
ingrowth and light availability is not yet captured correctly in most
models. Yet, several of the models that did feature an increase of
ingrowth with increasing basal area include a feedback between seed
production of mature trees and ingrowth. Thus, it seems that higher
light competition does not sufficiently compensate for increased seed
availability with higher basal area due to more mature trees.

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021 found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This may at
least partly explain e.g., the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

Based on the results of this study, it is clear that models of forest
dynamics need a focus on tree establishment modules to complement
earlier efforts to make them more robust regarding growth and mortality.
It remains uncertain what level of detail is required to model
establishment processes, and this must be addressed in future research.
We recommend that the further development of the establishment modules
is implemented as additional features that can be traced back, as done
here for the variants of ForClim and ForCEEPS, and that model complexity
and structure must always be connected with modeling objectives
[@albrich2020]. If it should be necessary to include more detail in the
establishment models, this will come with higher parameterisation
efforts. This will most likely lead to lower generalization because the
required data will have to be collected from specific locations, as
currently there is no general establishment dataset available.

### 2023-04-14 manuscript/supplementaryRW.Rmd
changes based on PICUS team feedback

```{r simObadult , echo=FALSE, fig.cap="Mean species richness at the stand level per site in the simulated versus observed species richness. The species richness is estimated using the Shannon index  calculated by basal area. A1) groups models that Simulated species composition in adults is higher than the observed, A2) groups models that Simulated species composition in adults is similar to the observed, A3) groups models where the imulated species composition in adults is lower than the observed. NOTE: The recruitment treshold recruitment for treshold 7 cm in the empirical data has several sites with no recruitment or no data and the Shannon index for those sites is NA", out.width='100%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "allAgesRichness.png"))
```

```{r simObr, echo=FALSE, fig.cap="Mean species richness per site in the simulated recruitment versus observed species richness. The species richness is estimated using the Shannon index is calculated by basal area.B1) groups models where the simulated species composition is closer than the other models to observed values, B2) groups models where the sites with lower species composition are underestimated, B3) groups models where the simulated species composition recruitment is lower in the whole range compare to the observe. NOTE: The recruitment treshold recruitment for treshold 7 cm in the empirical data has several sites with no recruitment or no data and the Shannon index for those sites is NA", out.width='100%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "recruitmentRichness.png"))
```

>>>>

```{r simObadult , echo=FALSE, fig.cap="Mean species richness at the stand level per site in the simulated versus observed species richness. The species richness is estimated using the Shannon index  calculated by basal area. A1) groups models where the simulated species composition in adults is higher than the observed, A2) groups models where the simulated species composition in adults is similar to the observed, A3) groups models where the simulated species composition in adults is lower than the observed. NOTE: The recruitment treshold recruitment for treshold 7 cm in the empirical data has several sites with no recruitment or no data and the Shannon index for those sites is NA", out.width='100%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "allAgesRichness.png"))
```

```{r simObr, echo=FALSE, fig.cap="Mean species richness per site in the simulated recruitment versus observed species richness in the recruitment. The species richness is estimated using the Shannon index using the basal area. B1) groups models where the simulated species composition in recruitment is higher than the observed, A2) groups models where the simulated species composition in recruitment is similar to the observed, A3) groups models where the simulated species composition in recruitment is lower than the observed. NOTE: The recruitment treshold recruitment for treshold 7 cm in the empirical data has several sites with no recruitment or no data and the Shannon index for those sites is NA", out.width='100%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "recruitmentRichness.png"))
```
